---
title: "hmi: hierarchical multiple imputation"
author: 
- name: "Matthias Speidel"
- name: "Jörg Drechsler"
- name: "Shahab Jolani"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hmi: hierarchical multiple imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---
references:
- id: Drechsler2015
  title: 'MI Double Feature: Multiple Imputation to Address Nonresponse and Rounding Errors in Income Questions'
  author:
  - family: Drechsler
    given: Jörg
  - family: Kiesl
    given: Hans
  - family: Speidel
    given: Matthias
  container-title: Austrian Journal of Statistics
  volume: 44
  URL: 'http://www.ajs.or.at/index.php/ajs/article/view/vol44-2-5'
  DOI:  10.17713/ajs.v44i2.77
  issue: 2
  publisher: 
  page: 
  type: article-journal
  issued:
    year: 2015
    month: 

- id: Hadfield2010
  title: 'MCMC Methods for Multi-Response Generalized Linear Mixed Models: The MCMCglmm R Package'
  author:
  - family: Hadfield
    given: Jarrod D.
  container-title: Journal of Statistical Software
  volume: 44
  URL: 'https://www.jstatsoft.org/article/view/v033i02'
  DOI: 10.18637/jss.v033.i02
  issue: 2
  publisher: 
  page: 
  type: article-journal
  issued:
    year: 2010
    month: 2

- id: Rubin1987
  title: 'Multiple Imputation for Nonresponse in Surveys'
  author:
  - family: Rubin
    given: Donald
  container-title: 
  volume: 
  URL: 
  DOI: 10.1002/9780470316696
  issue: 
  publisher: John Wiley \& Sons, Inc.
  page: 
  type: book
  issued:
    year: 1987
    month: 

- id: Wiencierz2012
  title: 'linLIR: linear Likelihood-based Imprecise Regression'
  author:
  - family: Wiencierz
    given: Andrea
  container-title: 
  volume: 
  URL: 'https://cran.r-project.org/package=linLIR' 
  DOI: 
  issue: 
  publisher: CRAN
  page: 
  type: article-journal
  issued:
    year: 2012
    month:
---


## Purpose of package
The hmi package allows the user to run single level and multilevel imputation models.
The big additional benefit of this package is the user-friendliness. 
It is designed for researchers, experienced in running single und multilevel analysis models, but not in writing own multilevel imputation routines.

The user just has to pass the data to the main function and, optionally, his analysis model.
Basically the package then translates this analysis model into commands to impute the data according to it with functions from `mice`, `MCMCglmm` or routines build for this package.

## Basic functionality

The main function that wraps up all sub functions is `hmi`. 

In the most simple case, the user just passes his `data` to `hmi`. In this case all variables with missing values are imputed based on a single level imputation model including the other variables. 
The situation, for which the package was built for, is that the user additionally passes his analysis model as `model_formula` to `hmi` (and defines more details if he wants to).
The function then analyzes the `model_formula`, checks whether it suits to the data given and runs some other checks on the data given (size of the data, number of remaining observations etc.). 

### Output of `hmi`
The package is build to be compatible with `mice`; especially with regard the output. `hmi` returns, like `mice`, a so-called `mids`-object (multiply imputed data set).

This allows the user, familiar with `mice` to use functions designed for `mice`-outputs without switching barriers. For example, running the generic `plot()`-function on a `mids`-object calls the function `plot.mids` showing the means and variances of the imputed variables over the different imputations, regardless whether the `mids`-object came from `mice` or `hmi`. Or he could call the `complete`-function delivered by `mice` to get a completed data set where the NAs are replaced by the imputed values.

## The different types of imputation routines / the supported types of variables
Different variable types require different imputation routines. For example for binary variables it would be unpleasant to impute other values than `0` and `1`. And factor variables with levels `"A"`, `"B"` and `"C"` need an imputation routine different to the ones for binary and continuous variables.

To determine which imputation routine shall be used, we first have to decide whether a single level or multilevel model shall be used. This decision is mainly based on the `model_formula` given by the user. The formula is decomposed into its fixed effects, random effects and cluster variable parts (if present). If the cluster variable and the random effect variables are actually present in the data set and available in the moment of imputation, a multilevel model is run. In all other cases (i.e. not available or not specified) a single level model is run.

The second question is which type the variable is of. We distinguish eight different types of variable. The next sections describe how we assign a type to a variable and how the imputation model works for these types. For some special cases the rules of assignment might give unwanted results. Therefore the user can specify the types of the variables in advance by setting up a `list_of_types`. Section *Pre-definition of the variable types* explains how this is done.

`MCMCglmm` assumes for each type of variable, that there is a latent variable $l$ present which can be expressed by fix and random effects. So $l = X \cdot \beta + Z \cdot u + \varepsilon$ [cf. @Hadfield2010 eq. 3]. The probability of observing $y_i$ is conditioned on $l_i$: $f_i(y_i|l_i)$, with $f_i$ being the probability density function (pdf) for $y_i$. More about the theory behind `MCMCglmm` can be found in the below.

For completeness: each imputation routine starts with some cleanup. This includes for example removing linear dependent variables (or other variables likely to hamper the imputation model like factors with more then 10 levels) from the current imputation.

### Binary variables (keyword `"binary"`)
Data are considered to be binary if there are only two unique values. This includes for example `0` and `1` or `"m"` and `"f"`. 

The single level imputation model is a logistic regression for a binomial family with a logit link. Based on this model new (Bayesian) imputation parameters are drawn. Those parameters are then used to sample binary observations, given the other covariates. This is implemented in the `mice.impute.logreg`-function which is called when running `mice` with the `method = "logreg"`.

In the multilevel model `MCMCglmm` is called with `family = categorical`. This uses the pdf $\exp(l)/(1+\exp(l))$

Settings where our rule of classification might fail are small data, or data with very few observed individuals or if a third possible category is unobserved. E.g. in a small health survey it could happen that none of the respondents reported having had two (or more) Bypass operations. So here a count variable would falsely be classified as binary.


### continuous variables (keyword `"cont"`)
Any numeric vector, that isn't one of the other types, is considered to be continuous.

In the single level model, `mice.impute.norm` from `mice` is called. This routine first draws imputation parameters (regression coefficients and residual variance) and then draws imputation values with these parameters. 

In the multilevel model `MCMCglmm` is called with `family = categorical`. This uses the normal distribution.

### semicontinuous variables (keyword `"semicont"`)
A continuous variable with more than 5\% values being 0, is defined being "semicontinuous".

The first step of imputing semicontinuous variables is to temporarily change internally all non-zero values to 1. Then via a binary imputation (based on the temporarily 0/1 variable) it is decided for the missing values whether they shall be 0 or non-zero. 

In a third step, for those being chosen to be non-zero, we run a continuous imputation model based on the originally non-zero observations. (Missing values, chosen to be 0, don't need further treatment, their imputation values is just 0). 

### rounded continuous variables (keyword `"roundedcont"`)
If more than 50% of the data are divisible by 5, they are considered to be "rounded continuous".
For example the income in surveys is often reported rounded by the respondents.

For this type of variable, we use our own imputation routine.

It estimates a model for the rounding degree G and for the variable Y itself, then parameters for the joint distribution of G and Y are drawn and afterward used to impute values. Not only missing values get a new imputed value, but also values with an interval response (e.g. "between 1500 and 2000") and (presumably) rounded responses.

Individuals with NAs get imputed values drawn from the normal distribution with the estimated parameters from the joint distribution.
Interval responses get imputed values drawn from the truncated normal distribution.
For individuals with (presumably) rounded responses, values are drawn for G and Y and then checked whether this combination could explain the actual observed value of Y for this observation. E.g. if 2950 is observed then the combination (G = degree 100, Y = 3000) would not fit to the observed response. In the case of a mismatch, the process is repeated until G and Y match.

The process is described in detail in [@Drechsler2015].

### interval variables (keyword `"interval"`)
We see interval data as a special case of imprecise observations given as a mathematical interval $[l;~u]$ with $l \leq u$. For example a person could refuse to report its precise income $y$, but is willing to report that it is something between 1500 and 2000. In this case the interval $[1500; ~2000]$ is the observed value for this individual. Precise answers like $3217$ can be seen as special cases of interval data where $l=u$, here $[3217;~3217]$. Missing values can be seen as the extreme case $[-\infty;~\infty]$.

To our knowledge, there is no standard in R for interval data. One possibility would be to generate *two* variables for the lower and the upper bounds of the data. Based on this approach [@Wiencierz2012] set up the `idf`-objects (interval data frame) in her package `linLIR`. We didn't follow this approach for our package because it would need an inconvenient workflow to link both variables appropriately. Instead, we define a new class `interval` for interval variables. Interval variables actually come in *one* variable. Technically one observation in such an interval variable is `"l;u"` with `l` (resp. `u`) being a scalar with optional decimal places in American notation (with a full stop. E.g. `"1234.56;3000"`) or `-Inf` (resp. `Inf`).

Within most `R` functions such an `interval`-variable will be treated as a factor. But it is a factor with maybe more than 100 categories. So we suggest not to use such a variable as covariate in a imputation model. Within `hmi` it would not be used as this would be too many categories. The main reason to use an `interval` variable is to impute this variable according to [@Drechsler2015].

We also implemented functions to run basic calculations on interval data (`+`, `-`, '*' and `/`), to generate interval data based on two vectors (`as_interval`) or to split interval data up into their lower and upper bounds (`split_interval`).

Furthermore, we want to encourage people working with interval data or variables and hope that a standard for this will emerge. For this reason we think, users should be able to switch easily between `idf` and `interval` objects as one might be better for one task and the other for a different task. So we implemented `idf2interval` and `interval2idf` which conveys an object from one format to the other (as far as possible). 

### count variables (keyword `"count"`)
Every vector with integers (which is not semicontinuous) is considered to be count data. By this definition, every continuous variable, rounded to the next integer is considered to be a count variable.

For both, single level and multilevel settings, we use `MCMCglmm` with the poisson distribution for the latent variable.

### categorical variables (keyword `"categorical"`)
Factor-variables (or variables with more than two categories - if they are not one of the previous types) are considered to be categorical variables. 

For the single level setting we use the `cart` approach in `mice`. This runs a regression tree for the observed data and then samples from suitable leaves for the individuals with missing values.

In the multi level setting, we use the `categorical` setting in `MCMCglmm` with runs a multilevel regression model for each category (based on the observed individuals). For the individuals with missing values, probabilities for each category are be calculated and than a category sampled based on these probabilities.  

### ordered categorical variables (keyword `"ordered_categorical"`)
In the special case, that a factor variable is ordered, we treat it as *ordered categorical*.

For the single level case `mice` is told to run an ordered logistic model.
`MCMCglmm` for the multilevel setting runs the `ordinal` model.

The assumption behind both models is that a latent variable $l$ is assumed to be present and dependent on how many thresholds $\gamma$, the variable exceeded, a higher category $k$ is observed.

### Intercept variable (keyword `"intercept"`)
A constant variable (only one kind of observation) is considered to be a intercept variable. 

## pre-definition of the variable types 

If you want to have manual control over the process which method is used for each variable, you can specify a `list_of_types`. This is a `list` where each list element has the name of a variable in the data frame. The elements have to contain a single character string denoting the type of the variable (the keywords from the previous section). With the function `list_of_types_maker`, the user can get the framework for this object.

In most scenarios this is shouldn't be necessary. One example where it might be necessary is when only two observations of a continuous variable are left - because in this case `get_type` interpret this variable to be binary. Or if you want to impute rounded continuous variables not as `"count"`, but as `"cont"`.
 
The example uses the data set `CO2` about the "Carbon Dioxide Uptake in Grass Plants", which comes along with `R`. If you run the common `str(CO2)`, you get the information that the variable `Plant` is an `Ord.factor w/ 12 levels`, `Type` is a `Factor w/ 2 levels`, `Treatment` is a `Factor w/ 2 levels`, `conc` is a `num` and so is `uptake`.

`hmi` draws similar conclusions. A difference would be that we call Factors with 2 levels "binary". Also the variable `conc` would be considered to be a special case of a continuous variable - a rounded continuous because every value of the ambient carbon dioxide concentration is divisible by at least 5.

You can see in advance how variables will be treated internally by `hmi` if you call the `list_of_types_maker`. For example `example_list_of_types <- list_of_types_maker(CO2)` gives you the following list:
```{r}
## $Plant
## [1] "ordered_categorical"
## 
## $Type
## [1] "binary"
## 
## $Treatment
## [1] "binary"
## 
## $conc
## [1] "roundedcont"
## 
## $uptake
## [1] "cont"
```

Now you can modify `example_list_of_types` according to your preferences. For example if you want the variable `conc` to be continuous, you can write `example_list_of_types[["conc"]] <- "cont"`. If you finished your modification on `example_list_of_types`, pass this list to `hmi` via its parameter `list_of_types`. In our example it would be `hmi(data = CO2, list_of_types = example_list_of_types)`. (Note, that `CO2` doesn't contain any missing value, so there is no need for imputation.)

## how to use it
To illustrate the use of `hmi`, we stick to the `CO2` data set; but as it has no missing values, we add those artificially:

```{r echo = TRUE}
library("lme4")
ex <- sleepstudy
set.seed(1)
ex[sample(1:nrow(CO2), size = 20), "Reaction"] <- NA
head(ex) # e.g. in line 5 there is a NA now.
```

A big part of the packages contribution is the multilevel imputation. For example, if your interest lies in modeling the effect of the carbon dioxide uptake rates (target variable `uptake`) by the ambient carbon dioxide concentration (explanatory variable `conc`) - and an intercept. Assuming that the effects of `conc` and the intercept on `uptake` can differ across the different plants (random effect variable `Plant`), your analysis model using `lmer` from the `lme4` package would be `lmer(uptake ~ 1 + conc + (1 + conc | Plant), data = example)`. Just for clarification how this model is read by `hmi`: the word left to `~` denotes the target variable, the parts right to `~` denote the fixed and random effects variables and the cluster ID. And in more detail, the parts within the parentheses left to `|` denote the random effects variables and the word right to `|` denotes the cluster ID:
```{r eval = TRUE, message = TRUE, cache = TRUE}
library("lme4")
lmer(formula = Reaction~1+Days+(1+Days|Subject), data = ex, na.action = na.omit)
```


If we would run `hmi(ex)` without further specification, we would end up in running a single level imputation. To make `hmi` run a multilevel imputation model, you have to specify a multilevel analysis model and this has two mandatory elements: 1. variables with a cluster specific effect (*random effects variables*) and 2. a variable indicating the clusters. By passing your analysis model formula to `hmi` you implicitly specify your imputation model(s). If there are more variables with missing values, then the other variables are tried to be imputed with a similar model. This means that maybe only one covariate in the analysis model becomes the target variable in the imputation model, but the random effects variables and the cluster ID keep the same (except in the case a random effects variable is to be imputed. In this case this variable is dropped from the random effects part of the imputation model). So here a multilevel imputation would be set up by (as we have only one variable to impute, we can set maxit to 1):

```{r eval = TRUE, echo = FALSE, results = "hide", message = FALSE, cache = TRUE}
library("hmi")
set.seed(1)
result_multi <- hmi(data = ex, model_formula = Reaction ~ 1 + Days + (1 + Days | Subject), maxit = 1, nitt = 1200, burnin = 200)
```

Now the imputation is complete and with the resulting `mids`-object you could do all the things described in the section above. 

Here we were especially interested in multilevel models, so we want to run our analysis model on the imputed data. By this it is meant that the model is run on every of the `M` completed data set and then the results are combined according to Rubin's combining rules [@Rubin1987].

`mice` has the functions `fit` and `pool` to do this. But only certain parameters of your model are pooled. 
```{r eval = TRUE, echo = TRUE, message = FALSE, cache = TRUE}
result_multi$pooling
```

# References
